**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach

### 2020-12-04 manuscript/ecosphere_resubmit_2/manuscript.Rmd
revisions submitted

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?  Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends. We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].

@@

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.

@@

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].

>>>>

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?  Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends. We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].

@@

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.

@@

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].
